Parallel global optimization on GPU \thanks{The research is supported by the grant of the Ministry of education and science of the Russian Federation (the agreement of August 27, 2013 No 02.B.49.21.0003)}

### 2019-10-07 2016/JoGO_GPU/BarkalovGergel.tex
JOGO2016 paper next version

This present paper considers parallel algorithms for the solution of problems of multiextremal optimization. In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of a priori information about the function, allowing us to link possible values of an optimizable function with known values in search points. Very often, such information about the solvable problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$. The objective function can be represented by the algorithm of calculation of its values in the search domain, i.e. it can be a ``black-box'' function. Similar problems are widespread in applications (problems of optimal design of objects and technological processes in various technical domains, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

@@

One of the promising directions of development in the domain of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years, and has become more programmable. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

@@

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} contains a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; comparison of the sequential algorithm and parallel CPU and GPU versions for the solution of a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within hyper interval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

Let us assume that function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,\; 0<L<\infty.
\]

There is a number of ways to adapt effective one-dimensional algorithms for solution of multidimensional problems; see, for example, diagonal partitions method in \cite{RefSergeyev2006} or simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}. Here we will note that a numerically constructed curve is an approximation to a theoretical Peano curve with accuracy $2^{-m}$ , where $m$ is a curve construction parameter.

Use of mapping of this kind allows a reduction of a multidimensional problem~(\ref{eq:1}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if function $\varphi(y)$ in domain $D$ satisfies Lipschitz condition, then function $\varphi(y(x))$ on interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where H{\"o}lder constant $H$ is linked to Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying H{\"o}lder condition.

The considered algorithm of the solution of this problem (here, according to \cite{RefStrongin2000}) presupposes construction of sequence of points $x^k$, where values of minimized function $z^k = f(x^k)$ are calculated. Let us call the process of calculation of the value of function (including construction of an image $y^k=y(x^k)$) ``trial'', and pair $(x^k, z^k)$ --  the ``trial result''. Set of pairs $\left\{(x^k, z^k)\right\}, 1\leq k\leq n,$ make up the search data collected using the method after carrying out $n$ steps. The rules that determine the work of the global search algorithm are as follows.

@@

Step 5. Carry out a trial in point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulas
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, t=1, t=k+1,
\]
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, 1<t<k+1.
\]

@@

Step 5. Carry out new trials in points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using formulas
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, t_j=1, t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, 1<t_j<k+1.
\]

@@

With respect to methods of global optimization, the operation which can be implemented effectively on GPU is simultaneous parallel computing of many values of the objective function. Naturally, for this purpose the function value calculation procedure has to be implemented on GPU. Transfers of information from CPU to GPU will be minimal: it is necessary only to transfer to GPU coordinates of trial points and to receive values of the function in these points. The functions that define processing of results of trials according to algorithm and require work with a large volume of collected search information can be effectively implemented on CPU.

A general scheme of organization of calculations using GPU is shown in Fig.~\ref{fig:3}. According to this scheme steps 1 -- 4 of a parallel global search algorithm are carried out on a CPU. The coordinates of $p$ points of trials calculated on step 4 of the algorithm are collected in an intermediate buffer, and then are transferred to the graphics processing unit. Calculation of function values in these points is carried out on GPU, and then the results of trials (again via an intermediate buffer) are transferred to CPU.

@@

The use of several hundred of parallel computing elements (as in modern GPU) might not give a speedup of the search process by hundreds of times. In this case, conditions of the theorem of non-redundant parallelization can be violated: the number of local extremums will be less than number of computing cores. Then (according to the theorem and its corollary) a parallel GSA will generate redundant trial points. Nevertheless, despite certain redundancy, the general operating time of a GPU algorithm will be less than the operating time of a CPU algorithm. This has been verified by computing experiments, the results of which are given in section \ref{sec:6}.

@@

It is noteworthy that the important parameter for starting any GPU program (unlike a parallel program on CPU) is not only the number of threads $p$, but also the size of the \textit{thread block}. Separate threads on GPU are grouped in blocks of threads of identical size, thus each thread block is executed on a separate multiprocessor. Threads in a thread block can interact effectively using information in shared memory and synchronization. For the purpose of assessment of the importance of the size of a thread block during the period of execution of a GPU algorithm let us perform a preliminary series of experiments with test function (\ref{eq:33}) with $N=5$. Let us vary number $p$ of used GPU cores and size $pB$ of a thread block. All the other parameters of the method do not change ($r=2.3$, $\epsilon = 10^{-3}$, evolvent construction parameter $m=10$). Table \ref{tab:2} shows the GPU algorithm operating time depending on $p$ and $pB$.
\begin{table}
	\caption{GPU algorithm operating time depending on thread block size}
	\label{tab:2}
	\center
	\begin{tabular}{llllll}
		\hline\noalign{\smallskip}
%		 $p$ & $pB=16$ & $pB=32$ & $pB=64$ & $pB=128$ & $pB=256$  \\
		$p$ & \multicolumn{5}{l}{ \hfil $pB$ \hfil } \\
		\noalign{\smallskip} \cline{2-6} \noalign{\smallskip}
		 & 16 & 32 & 64 & 128 & 256  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		100 &	39.43 &	41.5 &	39.71 &	39.85 &	39.61 \\
		300 &	24.84 &	24.52 &	24.54 &	24.66 &	25.08 \\
		500 &	9.25 &	9.45 &	9.24 &	9.4 &	9.3\\
		700 &	11.8 &	13.43 &	11.78 &	11.91 &	12.03\\
		900 &	10.9 &	10.91 &	10.85 &	10.86 &	10.96\\
		1000 &	10.92 &	10.8 &	10.76 &	10.9 &	10.74\\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

As is seen from this table, there is a speedup of algorithm in the case of an increase in the number of trials at one iteration up to the number of cores on GPU, with subsequent saturation. Owing to a weak information dependence the size of thread block $pB$ does not have determining influence, therefore let us use $pB=64$ in all subsequent experiments.

After fixing of parameters of the GPU let us make the main computing experiments on a series of multidimensional multiextremal problems. In work \cite{RefGaviano} a GKLS generator allowing generation of multiextremal optimization problems with known properties (number of local minimums, size of their domains of attraction, global minimum point, value of function, etc.) is described.

The results of numerical comparison of three sequential algorithms -- DIRECT \cite{RefJones}, DIRECT\textit{l} \cite{RefGablonsky} and global search algorithm (GSA) from section \ref{sec:2} -- are provided below (results of work of the first two algorithms are given in paper \cite{RefGaviano}). Numerical comparison was carried out on function classes \textit{Simple} and \textit{Hard} of dimension 4 and 5 from \cite{RefGaviano} since the solution of problems of dimension 2 and 3 demands a small number of iterations and use of GPU for solution of these problems is impractical. Global minimum $y^\ast$ was considered as found, if the algorithm generated trial point $y^k$ in $\delta$-vicinity of the global minimum, i.e. $\left\|y^k-y^\ast\right\|\leq\delta$. The size of the vicinity was selected (according to \cite{RefGaviano}) as $\delta = \left\|b-a\right\|\sqrt[N]{\Delta}$, $N$ -- problem dimension, $a$ and $b$ -- borders of search domain $D$, parameter $\Delta=10^{-6}$ at $N=4$ and $\Delta=10^{-7}$ at $N=5$. When using the GSA method for class \textit{Simple} parameter $r=4.5$ was selected, for class \textit{Hard} $r=5.6$ was selected; Peano curve construction parameter was fixed as $m=10$. The maximum allowable number of iterations was $K_{max} = 10^6$.

The average number of iterations $k_{av}$ performed by the method for solution of a series of problems from these classes is shown in table \ref{tab:3}. Symbol $>$ reflects a situation in which not all problems of a class were solved by a method. This means that the algorithm was stopped as the maximum allowable number of iterations $K_{max}$ was achieved. In this case, value $K_{max}=10^6$ was used for calculation of the average value of number of iterations $k_{av}$ that corresponds to the lower estimate of this average value. The number of unsolved problems is specified in brackets.

@@

Let us estimate now the speedup when using a parallel global search algorithm implemented on CPU depending on number $p$ of used cores. Table \ref{tab:4} shows the average number of iterations performed by the method during solution of a series of problems, and table \ref{tab:5} shows the average time of solution of one problem. Tables \ref{tab:6} and \ref{tab:7} show speedup on time $S(p)$ and iterations $s(p)$ respectively; speedup of parallel algorithm was measured in relation to the sequential one ($p=1$).

>>>>

This present paper considers parallel algorithms for the solution of problems of multiextremal optimization. In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of a priori information about the function, allowing us to link possible values of an optimizable function with known values in search points. Very often, such information about the solvable problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$. The objective function can be represented by the algorithm of calculation of its values in the search domain, i.e. it can be a ``black-box'' function. Similar problems are widespread in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

@@

One of the promising directions of development in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years, and it became more programmable. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

@@

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} contains a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize the speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of the parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; a comparison of the sequential algorithm and parallel CPU and GPU versions for solving a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within a hyper interval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

Let us assume that the function $\varphi$ satisfies a Lipschitz condition with a priori unknown constant $L$
\[
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D,\; 0<L<\infty.
\]

There is a number of ways to adapt effective one-dimensional algorithms for solving multidimensional problems; see, for example, the diagonal partitions method in \cite{RefSergeyev2006} or the simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}. Here we will note that a numerically constructed curve is an approximation to a theoretical Peano curve with an accuracy $2^{-m}$ , where $m$ is a curve construction parameter.

By using mapping of this kind it is possible to reduce the multidimensional problem~(\ref{eq:1}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, then the function $\varphi(y(x))$ on the interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is linked to the Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.

The considered algorithm for solving this problem (here, according to \cite{RefStrongin2000}) assumes constructing a sequence of points $x^k$, where the values of minimized function $z^k = f(x^k)$ are calculated. Let us call the process of calculating the function value (including the construction of an image $y^k=y(x^k)$) the ``trial'', and the pair $(x^k, z^k)$ --  the ``trial result''. Set of pairs $\left\{(x^k, z^k)\right\}, 1\leq k\leq n,$ make up the search data collected using the method after carrying out $n$ steps. The rules that determine the work of the global search algorithm are as follows.

@@

Step 5. Carry out a trial in point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulas
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, \textrm{ if } t=1 \textrm{ or } t=k+1,
\]
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t<k+1.
\]

@@

Step 5. Carry out new trials in points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using formulas
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, \textrm{ if } t_j=1 \textrm{ or } t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t_j<k+1.
\]

@@

With respect to methods of global optimization, the operation which can be implemented effectively on GPU is simultaneous parallel computing of many values of the objective function. Naturally, for this purpose the function value calculation procedure has to be implemented on GPU. Transfers of information from CPU to GPU will be minimal: it is necessary only to transfer to GPU coordinates of trial points and to receive values of the function in these points. The functions that determine the processing of trial results according to algorithm and require work with a large volume of collected search information can be effectively implemented on CPU.

A general scheme of organization of calculations using GPU is shown in Fig.~\ref{fig:3}. According to this scheme steps 1 -- 4 of a parallel global search algorithm are carried out on a CPU. The coordinates of $p$ points of trials calculated at step 4 of the algorithm are collected in an intermediate buffer, and then are transferred to the graphics processing unit. Calculation of function values in these points is carried out on GPU, and then the results of trials (again via the intermediate buffer) are transferred to CPU.

@@

The use of several hundred of parallel computing elements (as in modern GPU) might not give a speedup of the search process by hundreds of times. In this case, conditions of the theorem of non-redundant parallelization can be violated: the number of local extremums will be less than the number of computing cores. Then (according to the theorem and its corollary) a parallel GSA will generate redundant trial points. Nevertheless, despite certain redundancy, the general operating time of a GPU algorithm will be less than the operating time of a CPU algorithm. This has been verified by computing experiments, the results of which are given in section \ref{sec:6}.

@@

It is noteworthy that an important parameter for starting any GPU program (unlike a parallel program on CPU) is not only the number of threads $p$, but also the size of the \textit{thread block}. Separate threads on GPU are grouped in blocks of threads of identical size, thus each thread block is executed on a separate multiprocessor. Threads in a thread block can interact effectively using information in shared memory and synchronization. For the purpose of assessing the importance of the size of a thread block during the period of execution of a GPU algorithm let us perform a preliminary series of experiments with the test function (\ref{eq:33}) with $N=5$. Let us vary the number $p$ of the used GPU cores and the size $pB$ of a thread block. All the other parameters of the method do not change ($r=2.3$, $\epsilon = 10^{-3}$, the evolvent construction parameter $m=10$). Table \ref{tab:2} shows the GPU algorithm operating time depending on $p$ and $pB$.
\begin{table}
	\caption{GPU algorithm operating time depending on thread block size}
	\label{tab:2}
	\center
	\begin{tabular}{llllll}
		\hline\noalign{\smallskip}
%		 $p$ & $pB=16$ & $pB=32$ & $pB=64$ & $pB=128$ & $pB=256$  \\
		$p$ & \multicolumn{5}{l}{ \hfil $pB$ \hfil } \\
		\noalign{\smallskip} \cline{2-6} \noalign{\smallskip}
		 & 16 & 32 & 64 & 128 & 256  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		100 &	39.43 &	41.5 &	39.71 &	39.85 &	39.61 \\
		300 &	24.84 &	24.52 &	24.54 &	24.66 &	25.08 \\
		500 &	9.25 &	9.45 &	9.24 &	9.4 &	9.3\\
		700 &	11.8 &	13.43 &	11.78 &	11.91 &	12.03\\
		900 &	10.9 &	10.91 &	10.85 &	10.86 &	10.96\\
		1000 &	10.92 &	10.8 &	10.76 &	10.9 &	10.74\\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

As is seen from this table, there is a speedup of the algorithm in the case of an increase in the number of trials at one iteration up to the number of cores on GPU, with subsequent saturation. Owing to a weak information dependence the size of the thread block $pB$ does not have determining influence, therefore let us use $pB=64$ in all subsequent experiments.

After fixing the GPU parameters let us make the main computing experiments on a series of multidimensional multiextremal problems. In work \cite{RefGaviano} a GKLS generator allowing generation of multiextremal optimization problems with known properties (number of local minima, size of their domains of attraction, global minimum point, function value, etc.) is described.

The results of numerical comparison of the three sequential algorithms -- DIRECT \cite{RefJones}, DIRECT\textit{l} \cite{RefGablonsky} and global search algorithm (GSA) from section \ref{sec:2} -- are provided below (results of work of the first two algorithms are given in paper \cite{RefGaviano}). Numerical comparison was carried out on function classes \textit{Simple} and \textit{Hard} of dimension 4 and 5 from \cite{RefGaviano} since the solution of problems of dimension 2 and 3 demands a small number of iterations and it is impractical to use GPU for solving these problems. Global minimum $y^\ast$ was considered to be found, if the algorithm generated trial point $y^k$ in $\delta$-vicinity of the global minimum, i.e. $\left\|y^k-y^\ast\right\|\leq\delta$. The size of the vicinity was selected (according to \cite{RefGaviano}) as $\delta = \left\|b-a\right\|\sqrt[N]{\Delta}$, $N$ -- problem dimension, $a$ and $b$ -- borders of search domain $D$, parameter $\Delta=10^{-6}$ at $N=4$ and $\Delta=10^{-7}$ at $N=5$. When using the GSA for the class \textit{Simple} parameter $r=4.5$ was selected, for the class \textit{Hard} $r=5.6$ was selected; Peano curve construction parameter was fixed as $m=10$. The maximum allowable number of iterations was $K_{max} = 10^6$.

The average number of iterations $k_{av}$ performed by the method for solving a series of problems from these classes is shown in table \ref{tab:3}. Symbol $>$ reflects a situation in which not all problems of the class were solved by the method. This means that the algorithm was stopped as the maximum allowable number of iterations $K_{max}$ was achieved. In this case, value $K_{max}=10^6$ was used for calculating the average value of the number of iterations $k_{av}$ that corresponds to the lower estimate of this average value. The number of unsolved problems is specified in brackets.

@@

Let us estimate now the speedup when using the parallel global search algorithm implemented on CPU depending on the number $p$ of the used cores. Table \ref{tab:4} shows the average number of iterations performed by the method when solving a series of problems, and table \ref{tab:5} shows the average time requared to solve one problem. Tables \ref{tab:6} and \ref{tab:7} show speedup in time $S(p)$ and iterations $s(p)$ respectively; speedup of the parallel algorithm was measured in relation to the sequential one ($p=1$).
